---
title: "Final_Fall2015_TValentine.Rmd"
author: "Thomas K. Valentine"
date: "December 10, 2015"
output:
  html_document:
    highlight: pygments
    theme: cerulean
widgets     : [mathjax]            # {mathjax, quiz, bootstrap}
---


# Final Exam

**“I affirm that I have received no support in taking this exam from any individual other than the Professor for this course”. - TV**

## Instructions

Please **email** to me all of the materials necessary for another person to run your R Markdown file, including:
  - The R project (`.Rproj`) file
	- The R Markdown document (`.Rmd`) of your analyses
	- An HTML document (`.html`) compiled from your R Markdown document.
	- Any data or other files neede to run the analyses in your R Markdown document.

# Problems 

## Set 1
The first set of questions pertain to Olympic gold medal-winning 100 meter sprint times for men and women.
```{r eval=TRUE}
sprinters <- read.csv("input/sprinters.csv")
```

| Variable | Description |
|:--|:--|
| `year` |    Year of race |
| `finish` |     completion time in seconds |
| `women` |    indicator for female or male time |

(@) (4 points) Estimate the model
    $$
    \begin{aligned}[t]
    \mathtt{finish}_i &= \beta_0 + \beta_1 \mathtt{year}_i + \beta_2 \mathtt{women}_i + \beta_3 \mathtt{women}_i \times \mathtt{year}_i  + \epsilon_i
    \end{aligned}
    $$

(@) (4 points) Given your results for the model above, what is the marginal effect of `year` for women? For men? Interpret each in a sentence. 


(@) (4 points) Create a plot of the fitted values (with confidence intervals) with respect to `year`, grouped by `women`. Do not use `geom_smooth()`.


(@) (2 points) Create the same plot as in the previous problem for the model
    $$
    \begin{aligned}[t]
    \log (\mathtt{finish}_i) &= \beta_0 + \beta_1 \mathtt{year}_i + \beta_2 \mathtt{women}_i + \beta_3 \mathtt{year}_i \times \mathtt{women}_i + \epsilon_i
    \end{aligned}
    $$


(@) (4 points) Explain in no more than 2 sentences how this specification differs from the prior specification.

(@) (6 points) Rerun the analysis and recreate the plot for the model:
    $$
    \begin{aligned}[t]
    \mathtt{Finish}_i &= \beta_0 + \beta_1 \mathtt{year}_i + \beta_2 \mathtt{women}_i + \beta_3 \mathtt{year}_i × \mathtt{women}_i \\
	& + \beta_4 \mathtt{year}_i^2 + \beta_5 \mathtt{year}_i^2 × \mathtt{women}_i + \epsilon_i
    \end{aligned}
    $$


(@) (4 points) Explain in 2-3 sentences how this specification differs from the previous models. 


(@) (4 points) Compare the visual fit of these models to the data within the observed period. Which (if any) do you find to be plausible fits?

(@) (6 points) Do these models have different predictions for the Olympics of 2156? Why or why not?


(@) (6 points) Now create a new variable, the ratio of men’s time to women’s time in each year. Log-transform this variable and regress it on year. Plot the results, with confidence intervals, on the scale of the ratio men’s time to women’s time (i.e., transform it back from the natural logarithm).

##Set 2

For this problem, we will use a cleaned-up version of the dataset of Michael Ross, "Does Oil Hinder Democracy?" *World Politics*, 2001.That paper estimated a time series cross-section model of Polity scores regressed on oil exports and a battery of controls. In this problem, we will focus on a single cross-section, and instead focus on model fitting.

Load and pre-process this data using the following code:
```{r message = FALSE}
library(dplyr)
rossoil <- read.csv("input/rossoildata.csv") %>%
   arrange(id1, year) %>%
   group_by(id1) %>%
   mutate(oilL5 = lag(wdr123, 5) / 100,
          metalL5 = lag(wdr313, 5) / 100,
          GDPpcL5 = lag(wdr135, 5) / 100,
          islam = islam / 100) %>%
   filter(year == 1990) %>%
   select(regime1, oilL5, metalL5, GDPpcL5, islam, oecd, cty_name, id, id1) %>%
   na.omit() %>%
   ungroup()
```

The above code:

- `lag()` calculates the lag values of `oil`, `metal`, and `GDPcap`. The second argument of `5`, means that it calculates a lag of 5.[^1]
- `group_by` ensures that the lag values are only calculated within each country.
- Keeps only observations from 1995 and a subset of variables
- Omits missing values with `na.omit()`.
- `ungroup()` ensures that the data is no longer grouped by `id`.
  If you try to use `summarize()` while the data is still grouped by `id`, you would
  not get the results that you thought you would.

The variables are: 

| Variable | Description |
|:--|:--|
| `regime1` |    1–10 scale increasing in democracy; computed from Polity components |
| `oilL5` |      Fuel exports as a proportion of GDP, lagged 5 years |
| `metalL5` |    Ore and mineral exports as a proportion of GDP, lagged 5 years |
| `GDPpcL5` |    per capita GDP in PPP dollars, lagged 5 years |
| `islam` |      Muslims as a proportion of population, 1970 data |
| `oecd` |       Dummy for rich industrialized countries |
| `cty_name` |   The name of the country observed |
| `id` |         A three character abbreviation of the country name |
| `id1` |        A numeric country code |


(@) (4 points) Estimate a linear regression of `regime1` on `oilL5`, `metalL5`, `GDPpcL5`, `islam`, and `oecd`.

(@) (8 points) Provide a substantive interpetation of each coefficient and note whether a given coefficient is statistically significant. 


(@) (4 points) Calculate the expected change in `regime1` given a change in `oilL5` from the 50th percentile to the 95th percentile of the fully observed data, all else equal.

(@) (8 points) Using the residuals from the regression in above, create the following diagnostic plots:
    
    - Plot the standardized residuals against the fitted values
  
    - Plot the standardized residuals against each covariate
  
    - Plot $Y$ versus $\hat{Y}$

(@) (6 points) What do these diagnostics tell you about the presence of heteroskedasticity, specification error, and outliers?

(@) (4 points) Rerun the regression using any 2 non-linear transformations for any of the covariates ("x variables") based upon your theoretical expectations and/or your diagnostics above). You're welcome to explore several specifications, but only present the specification that you think is best and describe why you choose to transform what you did. You WILL NOT be graded on whether the model is ultimatly a good fitting model.

(@) (4 points) For your new model, record and compare the adjusted $R^2$, AIC score, and BIC score for your new model.

(@) (4 points) For both of your new specifications, compute the expected change in `regime1` given a change in `oilL5` from the 50th percentile to the 95th percentile of the fully observed data.

(@) (4 points) Assess how much substantive difference finding the best model makes. Be specific and concrete; i.e., show what each model does. 
  
(@) (4 points) What other problems in the specification or estimation method remain unaddressed by our efforts (no more than 2-3 sentences)?

(@) Try a new specification that alters the dependent variable in order to address the heteroskedasticity problem. Assess whether your new model successfully addresses the problem. 

